Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-09-13 analysis/paper/paper.qmd
Updated manuscript

In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden. X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of, what has been classified by archaeologists as, brecciated quartz and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and ED-XRF (@fig-artefacts; @fig-artefact-size).

@@

The dataset includes two main groups of petrographic material, as classified by archaeologists: quartzite and brecciated quartz (Swe. "*brecciekvarts*") (@fig-artefacts; @tbl-point-data). Geological definitions of quartzite include characteristics such as the material being exceptionally hard, predominantly consisting of quartz, and with conchoidal fracturing [@skolnick65; @howard05; @prieto_etal19]. A reliable classification would thus be based on a combination of chemical and physical (macroscopic and microscopic) properties. Archaeologists, however, tend to classify the material in the field on the basis of visual characteristics and perceived hardness. This can lead to considerable ambiguity, but in general, a material will be archaeologically classified as quartzite in Northern Sweden if it is an exceptionally hard, fine-grained, quartz-rich rock.

@@

This study includes `r length(quartzite)` and `r length(bquartz)` brecciated quartz samples. The inclusion of brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated. The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67]. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. No detailed information on these microscopic features were mentioned in his original publication. Brecciated quartz in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g. @holm91: 24] (@fig-artefacts). As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

@@

### X-Ray Fluorescence (XRF) {-}
XRF spectroscopy is a well-established method for petrographic characterization in archaeology [@hall60; @shackley11a]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample [@sitko_zawisa12: 143-159]. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated, as was done in this project.

Measurements were collected using the Energy Dispersive (ED) Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand. The reference calibration used for element quantification was mining mode [@knight_etal21: 2]. The integration time was set to 130 seconds. Higher spectral resolution can be achieved if used with a Wavelength Dispersive (WD) XRF system [@menne_etal20]. However, these are limited to lab-based environments and not applicable in the field, which is central to the current study.

@@

The PCA model (@fig-raman-pca) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks. Samples where the D/G peaks dominate in terms of intensity cluster towards the positive side of PC 1, with those that completely lack or have weaker D/G bands on the opposite side. Similarly, along PC 2 (`r raman.pc2var` %) the samples that have higher intensity D/G peaks gravitate towards the top of the score plot. Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot. One of these is a brecciated quartz sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material. The second sample, however, stands out from the rest of the material. It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals. It does not fit with the “expected” image of a quartzite and may be a case of erroneous archaeological classification. The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.

@@

The XRF spectrometer is not capable of detecting all the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample. The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18]. The analysis shows that ~14~Si makes up the main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot). The majority of the readings resulted in ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. Whilst this could be due to the limitations of the instrumentation, it could also be a result of the sampling. As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%. ~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn) recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.

@@

Cluster 2 is distinguished by a higher titanium (~22~Ti), ~26~Fe and ~40~Zr content. It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue. ~13~Al and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters. As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism. Cluster 3 features a more limited range of values between the samples and includes all the brecciated quartz.

>>>>

In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden (https://samlingar.vbm.se/). X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of, what has been classified by archaeologists as, *brecciated quartz* and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and ED-XRF (@fig-artefacts; @fig-artefact-size).

@@

The dataset includes two main groups of petrographic material, as classified by archaeologists: quartzite and *brecciated quartz* (Swe. "*brecciekvarts*") (@fig-artefacts; @tbl-point-data). Geological definitions of quartzite include characteristics such as the material being exceptionally hard, predominantly consisting of quartz, and with conchoidal fracturing [@skolnick65; @howard05; @prieto_etal19]. A reliable classification would thus be based on a combination of chemical and physical (macroscopic and microscopic) properties. Archaeologists, however, tend to classify the material in the field on the basis of visual characteristics and perceived hardness. This can lead to considerable ambiguity, but in general, a material will be archaeologically classified as quartzite in Northern Sweden if it is an exceptionally hard, fine-grained, quartz-rich rock.

@@

This study includes `r length(quartzite)` and `r length(bquartz)` *brecciated quartz* samples. The inclusion of *brecciated quartz* in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated. The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67]. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. No detailed information on these microscopic features were mentioned in his original publication. *Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g. @holm91: 24] (@fig-artefacts). As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

@@

### X-Ray Fluorescence (XRF) {-}
XRF spectroscopy is a well-established method for petrographic characterization in archaeology [@hall60; @shackley11a]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample [@sitko_zawisa12: 143-159]. Portable XRF instruments typically feature a set of filters to achieve this, calibrated for different mediums (e.g. soil, metals, minerals) [@knight_etal21: 2].

Measurements were collected using the Energy Dispersive (ED) Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand. The reference calibration used for element quantification was mining mode [@knight_etal21: 2; @thermo18]. The integration time was set to 130 seconds. Higher spectral resolution can be achieved if used with a Wavelength Dispersive (WD) XRF system [@menne_etal20]. However, these are limited to lab-based environments and not applicable in the field, which is central to the current study.

@@

The PCA model (@fig-raman-pca) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks. Samples where the D/G peaks dominate in terms of intensity cluster towards the positive side of PC 1, with those that completely lack or have weaker D/G bands on the opposite side. Similarly, along PC 2 (`r raman.pc2var` %) the samples that have higher intensity D/G peaks gravitate towards the top of the score plot. Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot. One of these is a *brecciated quartz* sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material. The second sample, however, stands out from the rest of the material. It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals. It does not fit with the “expected” image of a quartzite and may be a case of erroneous archaeological classification. The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.

@@

The XRF spectrometer is not capable of detecting all the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample. The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18]. The analysis shows that ~14~Si makes up the main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot). The majority of the readings resulted in ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. Whilst this could be due to the limitations of the instrumentation, it could also be a result of the sampling. As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%. Certain elements recorded a high number of values below the LOD (“missing values”), including ~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn). Although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.

@@

Cluster 2 is distinguished by a higher titanium (~22~Ti), ~26~Fe and ~40~Zr content. It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue. ~13~Al and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters. As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism. Cluster 3 features a more limited range of values between the samples and includes all the *brecciated quartz*.
